Metric Spaces with Expensive Distances

### 2019-04-21 paper/expensive-distances-main.tex
interim

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
a {$1+\eps$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

\myparagraph{Contributions}
%
We propose simple algorithms for spanner construction and approximate nearest neighbor search
and evaluate them theoretically and experimentally in the defined cost model.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested example. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

For approximate nearest neighbor, we devise a simple randomized incremental algorithm and show that
the number of distance queries to find an approximate nearest neighbor is $O(\log n)$ in expectation.
Our proof is based on the well-known observation that the nearest neighbor changes $O(\log n)$ times
in expectation when traversing the sequence of points, combined with a packing argument certifying that
only a constant number of distances needs to be computed in-between two minima.
Our experimental results match the theoretical predictions
%We also experimentally evaluate our approach and observe that the approach follows 
%roughly the theoretical prediction.

@@

We give the details of the \textsc{UpdateBounds} procedure next.
Suppose that
$\dist(p_i,p_j)=v\in\RR$ has been computed.
First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$, since the distance
of $p_i$ and $p_j$ is exactly $v$.
To update the upper bound of some entry $b_{k,\ell}$,
we observe that the shortest path from $p_k$ to $p_\ell$ might now
go through the new edge. Hence, we update
\[
    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
\]
Repeating this for all $k,\ell$ yields the updated upper bounds.
Note that this results in $O(n^2)$ arithmetic operations,
but no distance computation.

@@

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
%$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
%$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
%$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  0  &  13557\\
$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  0  &  9383\\
$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  0  &  5146\\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

\section{Approximate nearest neighbors}
\label{sec:ann}
We consider the standard problem of finding an approximate nearest neighbor: given
$n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. This notation will be fixed throughout this 
section, and we shall also use the shorthand notation
\[
    r_i := \dist(p_i, q).
\]
We assume for simplicity
that all exact pairwise distances $\dist(p_i, p_j)$ are already computed
(a slight modification of the algorithm can also be applied if only a spanner is available).
Our goal is to reduce the number of computed distances $\dist(p_i, q)$.

Our approach can be summarized as follows. 
Fix a random permutation of the points of $P$ 
and consider the points in that order 
(to simplify notation,
we re-index them, so the order is again $p_1, \dots, p_n$).
During the loop, we maintain lower bounds of each $p_i$
to the query point $q$, which are initially all set to $0$.
We also remember the closest neighbor $c$ that we have seen so
far and its distance $v$ to $q$.
We refer to the point $c$ as the \emph{candidate}.
We maintain the invariant that $c$ is an approximate nearest neighbor
to $q$ for the points $\{p_1,\ldots,p_i\}$.
When reaching the point $p_i$, we check whether the lower
bound $a_i$ satisfies $a_i\geq \frac{v}{1+\eps}$.
If so, $c$ remains an approximate nearest neighbor and
we do not query the distance of $p_i$ to $q$.
Otherwise, we compute $\dist(p_i,q)$ and update the lower bounds
of all points according to the newly computed distance.
If $p_i$ is closer to $q$ than $c$, we update $c$ and $v$
accordingly. At the end of the loop, $c$ is an approximate nearest
neighbor. The pseudocode of the procedure follows.

@@

We remark that we obtain an exact nearest neighbor algorithm
when setting $\eps$ to $0$, which means 
replacing the condition in the if-statement of the loop
with $a_i\geq v$.

@@

The full proof is given in the appendix \cref{sec:proof_ann}.

\section{Experiments on approximate nearest neighbors}
\label{sec:experiment_ann}

@@

\section{Additional Experimental Results on Spanners}
\label{sec:spanners_euclidean_addendum}
In this section we give additional plots that give more information
about the behaviour of \bgrdy spanner in Euclidean space.

@@

We shall need the following lemma, which is just a reformulation of the well-known
packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).

@@

Towards the proof, we will use the following geometric lemma which
can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
further distance computations of points very close to $p_i$ or very far from $p_i$
will be avoided.

@@

\definecolor{uququq}{rgb}{0.25,0.25,0.25}
\definecolor{xdxdff}{rgb}{0.49,0.49,1}
\definecolor{qqqqff}{rgb}{0,0,1}
\begin{figure}[ht!]
    \centering
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.5cm,y=0.5cm]
%\draw[->,color=black] (-1.5,0) -- (17,0);
%\foreach \x in {,2,4,6,8,10,12,14,16}
%\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
%\draw[->,color=black] (0,-7.5) -- (0,9.2);
%\foreach \y in {-6,-4,-2,2,4,6,8}
%\draw[shift={(0,\y)},color=black] (2pt,0pt) -- (-2pt,0pt) node[left] {\footnotesize $\y$};
%\draw[color=black] (0pt,-10pt) node[right] {\footnotesize $0$};
\clip(-1.5,-7.5) rectangle (17,9.2);
\draw [fill=black,fill opacity=0.5] (8.52,0.98) circle (1.02cm);
\draw [fill=black,fill opacity=0.15] (8.52,0.98) circle (4cm);
\draw [fill=black,fill opacity=0.95] (6.12,3.39) circle (0.42cm);
\draw [fill=black,fill opacity=0.5] (7.65,-3.14) circle (1.32cm);
\draw [line width=2pt] (3.51,0.63) circle (1.9cm);
\draw [dash pattern=on 2pt off 2pt] (3.51,0.63) circle (1.49cm);
\draw (3.51,0.63)-- (0.62,1.33);
\draw (3.51,0.63)-- (6.12,3.39);
\draw (8.52,0.98)-- (3.51,0.63);
\draw (8.52,0.98)-- (14.89,-3.85);
\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps)$};
\begin{scriptsize}
\fill [color=qqqqff] (3.51,0.63) circle (1.5pt);
\draw[color=qqqqff] (3.71,0.94) node {$q$};
\fill [color=xdxdff] (0.62,1.33) circle (1.5pt);
\draw[color=black] (3.24,1.41) node {$r_1/(1+\eps)$};
\fill [color=qqqqff] (8.52,0.98) circle (1.5pt);
\draw[color=qqqqff] (8.9,1.29) node {$p_2$};
\fill [color=uququq] (6.48,0.83) circle (1.5pt);
\fill [color=qqqqff] (7.83,2.21) circle (1.5pt);
\draw[color=qqqqff] (8.21,2.52) node {$p_3$};
\fill [color=qqqqff] (7.65,-3.14) circle (1.5pt);
\draw[color=qqqqff] (8.01,-2.82) node {$p_4$};
\fill [color=qqqqff] (6.62,0.6) circle (1.5pt);
\draw[color=qqqqff] (6.98,0.92) node {$p_5$};
\fill [color=qqqqff] (1.49,-2.39) circle (1.5pt);
\draw[color=qqqqff] (1.87,-2.08) node {$p_6$};
\fill [color=qqqqff] (1.95,-2.65) circle (1.5pt);
\draw[color=qqqqff] (2.31,-2.33) node {$p_7$};
\fill [color=qqqqff] (2.3,-2.75) circle (1.5pt);
\draw[color=qqqqff] (2.68,-2.43) node {$p_8$};
\fill [color=qqqqff] (2.79,-2.9) circle (1.5pt);
\draw[color=qqqqff] (3.17,-2.57) node {$p_9$};
\fill [color=xdxdff] (14.89,-3.85) circle (1.5pt);
\fill [color=uququq] (5.84,3.09) circle (1.5pt);
\fill [color=xdxdff] (4.52,4.29) circle (1.5pt);
\draw[color=xdxdff] (4.89,4.61) node {$p_1$};
\end{scriptsize}
\end{tikzpicture}
\caption{First two steps of the ANN algorithm. First $p_1$ is 
chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered at $p_2$, because their lower bounds allow us to discard them. Note that the point $p_5$, which
is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis. The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
\label{fig:ann_illustration}
\end{figure}

\begin{figure}[hb!]
    \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
    \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
    The exact nearest neighbor is the point $p_1$, next point $p_i$ is placed 
    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
    is the true nearest neighbor cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
    \label{fig:exact_bad_example}
\end{figure}

@@

@@

\begin{lemma}
\label{lem:sequence_lemma}
Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
\[\left(\frac{4(2+\eps) v_k}{\eps v_\ell}\right)^{d}\]
distances to $q$.
\end{lemma}
\begin{proof}
By the first part of \cref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
is computed, the second part of \cref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying \cref{lem:packing_lemma}.
\end{proof}

A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
the number of computed distances is a constant (since $v_k=v_\ell$). 
This means that to prove \cref{thm:ann_bound}, it would suffice to show
that the candidate changes only a logarithmic number of times in expectation.
While we have not found a simple proof for this claim, we can prove the statement with a slight variant of
that argument.

@@

Note that for $\eps>0$,
a minimum $p_k$ is not necessarily the candidate $c_k$ because a previous point
in the sequence close to $p_k$ might have caused the lower bound $a_k$ to be
in the interval $[\frac{v_k}{1+\eps},v_k]$, which leads to not
computing the distance $r_k$. However, it is true that
$v_k\leq (1+\eps)r_k$, because otherwise, $c_k$ would not be an approximate
nearest neighbor of $\{p_1,\ldots,p_k\}$.

@@

We point out that the proof fails for $\eps=0$ because in that case, we cannot exclude an $\eps$-ball
of close-by points as in the second part of \cref{lem:bound_lemma}, and the packing argument fails.
Indeed, as the example in \cref{fig:exact_bad_example} shows, there are point sets where the expected number
of distance computations for exact nearest neighbor is linear.

>>>>

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
a {$1+\eps$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbours
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

\myparagraph{Contributions}
%
We propose simple algorithms for spanner construction and approximate nearest neighbour search
and evaluate them theoretically and experimentally in the defined cost model.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested examples. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
outperforms naive computation of pairwise distances or runs slighly slower, depending on the approximation
quality.

For approximate nearest neighbour, we devise a simple randomized incremental algorithm and show that
the number of distance queries to find an approximate nearest neighbour is $O(\log n)$ in expectation.
Our proof is based on the well-known observation that the nearest neighbour changes $O(\log n)$ times
in expectation when traversing the sequence of points, combined with a packing argument certifying that
only a constant number of distances needs to be computed in-between two minima.
Our experimental results match the theoretical predictions.
%We also experimentally evaluate our approach and observe that the approach follows 
%roughly the theoretical prediction.

@@

We give the details of the \textsc{UpdateBounds} procedure next.
Suppose that
$\dist(p_i,p_j)=v\in\RR$ has been computed.
First, we reset $a_{i,j}$, $a_{j,i}$, $b_{i,j}$ and $b_{j,i}$ to $v$, since the distance
of $p_i$ and $p_j$ is exactly $v$.
To update the upper bound of some entry $b_{k,\ell}$,
we observe that the shortest path from $p_k$ to $p_\ell$ might now
go through the new edge. Hence, we update
\[
    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
\]
Repeating this for all $k,\ell$ yields the updated upper bounds.
Note that this results in $O(n^2)$ arithmetic operations,
but no distance computation.

@@

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
%$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
%$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
%$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
%$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  0  &  13557\\
$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  0  &  9383\\
$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  0  &  5146\\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

\section{Approximate nearest neighbours}
\label{sec:ann}
We consider the standard problem of finding an approximate nearest neighbour: given
$n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. This notation will be fixed throughout this 
section, and we shall also use the shorthand notation
\[
    r_i := \dist(p_i, q).
\]
We assume for simplicity
that all exact pairwise distances $\dist(p_i, p_j)$ are already computed
(a slight modification of the algorithm can also be applied if only a spanner is available).
Our goal is to reduce the number of computed distances $\dist(p_i, q)$.

Our approach can be summarized as follows. 
Fix a random permutation of the points of $P$ 
and consider the points in that order 
(to simplify notation,
we re-index them, so the order is again $p_1, \dots, p_n$).
During the loop, we maintain lower bounds of each $p_i$
to the query point $q$, which are initially all set to $0$.
We also remember the closest neighbour $c$ that we have seen so
far and its distance $v$ to $q$.
We refer to the point $c$ as the \emph{candidate}.
We maintain the invariant that $c$ is an approximate nearest neighbour
to $q$ for the points $\{p_1,\ldots,p_i\}$.
When reaching the point $p_i$, we check whether the lower
bound $a_i$ satisfies $a_i\geq \frac{v}{1+\eps}$.
If so, $c$ remains an approximate nearest neighbour and
we do not query the distance of $p_i$ to $q$.
Otherwise, we compute $\dist(p_i,q)$ and update the lower bounds
of all points according to the newly computed distance.
If $p_i$ is closer to $q$ than $c$, we update $c$ and $v$
accordingly. At the end of the loop, $c$ is an approximate nearest
neighbour. The pseudocode of the procedure follows.

@@

We remark that we obtain an exact nearest neighbour algorithm
when setting $\eps$ to $0$, which means 
replacing the condition in the if-statement of the loop
with $a_i\geq v$.

@@

The proof idea is the following. First, we notice
that current candidate $c$ allows us to exclude points
$p_i$ that are too far from $c$. This means that if $p_i$ is
outside of the ball of radius $2\dist(c, q)$ centered at $c$, then the lower
bound for $p_i$ will be larger than $\dist(c, q)$,
and discarding $p_i$ does not cost us a distance computation.
On the other hand, once we computed a distance to $c$
from a point $p_j$, which is inside this ball but happens to be farther from $q$ than $c$,
we can use it to discard points which are very close to $p_j$, 
namely, inside a small ball of radius approximately $\eps \dist(c, q)$, because we
are solving approximate nearest neighbour problem.
Then we use a standard packing argument for doubling spaces, applied to the large and small balls, to show
that the number of distance computations that we pay per candidate is constant.
It remains only to prove that the number of different candidates is $O(\log n)$
in expectation; there is a technicality that prevents using the standard
backwards analysis argument verbatim, but it can be adapted.
The full proof is given in \cref{sec:proof_ann}.

\section{Experiments on approximate nearest neighbours}
\label{sec:experiment_ann}

@@

\section{Additional Experimental Results on Spanners}
\label{sec:spanners_euclidean_addendum}
In this section we include plots that give more information
about the behaviour of \bgrdy spanner in Euclidean space.

@@

The following lemma is just a reformulation of the well-known
packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).

@@

The following geometric lemma
can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
further distance computations of points very close to $p_i$ or very far from $p_i$
will be avoided.

@@

\definecolor{uququq}{rgb}{0.25,0.25,0.25}
\definecolor{xdxdff}{rgb}{0.49,0.49,1}
\definecolor{qqqqff}{rgb}{0,0,1}
\begin{figure}[ht!]
    \centering
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.5cm,y=0.5cm]
%\draw[->,color=black] (-1.5,0) -- (17,0);
%\foreach \x in {,2,4,6,8,10,12,14,16}
%\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
%\draw[->,color=black] (0,-7.5) -- (0,9.2);
%\foreach \y in {-6,-4,-2,2,4,6,8}
%\draw[shift={(0,\y)},color=black] (2pt,0pt) -- (-2pt,0pt) node[left] {\footnotesize $\y$};
%\draw[color=black] (0pt,-10pt) node[right] {\footnotesize $0$};
\clip(-1.5,-7.5) rectangle (17,9.2);
\draw [fill=black,fill opacity=0.5] (8.52,0.98) circle (1.02cm);
\draw [fill=black,fill opacity=0.15] (8.52,0.98) circle (4cm);
\draw [fill=black,fill opacity=0.95] (6.12,3.39) circle (0.42cm);
\draw [fill=black,fill opacity=0.5] (7.65,-3.14) circle (1.32cm);
\draw [line width=2pt] (3.51,0.63) circle (1.9cm);
\draw [dash pattern=on 2pt off 2pt] (3.51,0.63) circle (1.49cm);
\draw (3.51,0.63)-- (0.62,1.33);
\draw (3.51,0.63)-- (6.12,3.39);
\draw (8.52,0.98)-- (3.51,0.63);
\draw (8.52,0.98)-- (14.89,-3.85);
\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps)$};
\begin{scriptsize}
\fill [color=qqqqff] (3.51,0.63) circle (1.5pt);
\draw[color=qqqqff] (3.71,0.94) node {$q$};
\fill [color=xdxdff] (0.62,1.33) circle (1.5pt);
\draw[color=black] (3.24,1.41) node {$r_1/(1+\eps)$};
\fill [color=qqqqff] (8.52,0.98) circle (1.5pt);
\draw[color=qqqqff] (8.9,1.29) node {$p_2$};
\fill [color=uququq] (6.48,0.83) circle (1.5pt);
\fill [color=qqqqff] (7.83,2.21) circle (1.5pt);
\draw[color=qqqqff] (8.21,2.52) node {$p_3$};
\fill [color=qqqqff] (7.65,-3.14) circle (1.5pt);
\draw[color=qqqqff] (8.01,-2.82) node {$p_4$};
\fill [color=qqqqff] (6.62,0.6) circle (1.5pt);
\draw[color=qqqqff] (6.98,0.92) node {$p_5$};
\fill [color=qqqqff] (1.49,-2.39) circle (1.5pt);
\draw[color=qqqqff] (1.87,-2.08) node {$p_6$};
\fill [color=qqqqff] (1.95,-2.65) circle (1.5pt);
\draw[color=qqqqff] (2.31,-2.33) node {$p_7$};
\fill [color=qqqqff] (2.3,-2.75) circle (1.5pt);
\draw[color=qqqqff] (2.68,-2.43) node {$p_8$};
\fill [color=qqqqff] (2.79,-2.9) circle (1.5pt);
\draw[color=qqqqff] (3.17,-2.57) node {$p_9$};
\fill [color=xdxdff] (14.89,-3.85) circle (1.5pt);
\fill [color=uququq] (5.84,3.09) circle (1.5pt);
\fill [color=xdxdff] (4.52,4.29) circle (1.5pt);
\draw[color=xdxdff] (4.89,4.61) node {$p_1$};
\end{scriptsize}
\end{tikzpicture}
\caption{First two steps of the ANN algorithm. 
%    First $p_1$ is 
%chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
%distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered 
%at $p_2$, because their lower bounds allow us to discard them. Note that the point $p_5$, which
%is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$
%in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$. 
%This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the
%brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis. 
The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
\label{fig:ann_illustration}
\end{figure}

\begin{figure}[hb!]
    \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
    \caption{Example of point set where exact nearest neighbour search cannot be accelerated by maintaining bounds.
    The exact nearest neighbour is the point $p_1$, next point $p_i$ is placed 
    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
    is the true nearest neighbour cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
    \label{fig:exact_bad_example}
\end{figure}

@@

In \cref{fig:ann_illustration} we illustrate this. First, $p_1$ is
chosen as the current candidate, and we must compute $\dist(p_2, q)$.
After that the algorithm will not compute
distance to any of the points inside the heavily shaded ball or outside
the lightly shaded ball that are centered at $p_2$,
because their lower bounds allow us to discard them.

@@

\begin{lemma}
\label{lem:sequence_lemma}
Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
\[\left(\frac{4(2+\eps) v_k}{\eps v_\ell}\right)^{d}\]
distances to $q$.
\end{lemma}
\begin{proof}
By the first part of \cref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
is computed, the second part of \cref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying \cref{lem:packing_lemma}.
See \cref{fig:ann_illustration}.
\end{proof}

A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
the number of computed distances is a constant (since $v_k=v_\ell$). 
This means that to prove \cref{thm:ann_bound}, it would suffice to show
that the candidate changes only a logarithmic number of times in expectation.
Let us consider \cref{fig:ann_illustration} again. Notice that the point $p_5$, which
is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points 
$p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are farther from 
$q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the 
candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, 
thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis,
and we need to modify it slightly.

@@

Note that for $\eps>0$,
a minimum $p_k$ is not necessarily the candidate $c_k$ because a previous point
in the sequence close to $p_k$ might have caused the lower bound $a_k$ to be
in the interval $[\frac{v_k}{1+\eps},v_k]$, which leads to not
computing the distance $r_k$. However, it is true that
$v_k\leq (1+\eps)r_k$, because otherwise, $c_k$ would not be an approximate
nearest neighbour of $\{p_1,\ldots,p_k\}$.

@@

We point out that the proof fails for $\eps=0$ because in that case, we cannot exclude an $\eps$-ball
of close-by points as in the second part of \cref{lem:bound_lemma}, and the packing argument fails.
Indeed, as the example in \cref{fig:exact_bad_example} shows, there are point sets where the expected number
of distance computations for exact nearest neighbour is linear.
